 

### 2021-04-17 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

**(paragraph on radiation as a driver)**
Incoming solar radiation is a strong driver of leaf temperature, relative to $T_{air}$, such that even at low maximum $T_{air}$, leaves exposed to high radiation loads can experience elevated $T_{leaf}$, greater than $T_{air}$. On the other hand, under conditions of constant radiation and water availability, increase in $T_{air}$ can induce evaporative cooling [Drake et al. 2018] or under high $T_{air}$, can cause leaf heat stress [Fauset et al. 2018]. Dense canopies that buffer understory tend to absorb and reflect incoming radiation, therefore maintain greater $T_{leaf}$ than $T_{air}$, whereas open-forests, with greater light interception to forest floors, in come cases, reflect a less pronounced gradient of leaf temperatures, and in other cases maintain greater $T_{leaf}$-$T_{air}$ in the lower canopies [**REF**]

@@

Particularly in dense forests, the upper, outer canopy leaves of taller trees are exposed to dramatically different biophysical conditions than leaves in the canopy interior or understory (Figs. `r fig_NEON_vertical`, `r SIfigs_NEON_all`). Exposed, typical sun leaves are often smaller in size and width, thicker, and hairier, with thicker cuticle and greater lobation than shade leaves, whereas shade leaves increase light capture efficiency by increasing their specific leaf area [SLA] and by investing in chlorophyll on a mass basis. "This enables shade leaves to realize similar light capture, under low light, to sun leaves at a lower biomass investment" [Chazdon et al. 1996, Poorter et al. 2000; @rozendaalPlasticityLeafTraits2006; Mathur et al. 2018]. Leaf traits are critical to maintaining desirable $T_{leaf}$ and, in turn, shaping leaf metabolism (following section) and carbon balance. It is unsuprising, then, that leaf traits vary dramatically across vertical gradients.

@@

Within-species, several studies point towards these common variations in leaf anatomical, structural and biochemical trait plasticity at a vertical intra-canopy gradient across individual trees [@sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020]. The differences among these traits contribute to leaf heat dissipation strategies and leaf temperature across the gradient [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016] Table `r table_leaf_traits` summarizes relevant traits for thermal sensitivity.

@@

Leaf temperature is generally assumed to increase with leaf area ($LA$) [ Wright et al. 2019; Tserej & Feeley 2021]. Yet, thermal imagery shows that effective leaf width is a better predictor of leaf temperature changes and leaf thermal time constant than leaf area[@leighInfluenceLeafSize2017]. With increases in effective leaf width, thermal time constant also increases due to increase in boundary layer thickness and resistance, along with stomatal resistance [@leighInfluenceLeafSize2017; Wright et al. 2019; @fauset_differences_2018]. Therefore, the rate of heat convection per unit area between leaf and air is lower for large leaves and greater for small leaves. Small leaves are able to dissipate heat more efficiently than large leaves [@leighInfluenceLeafSize2017]. Evidence shows that leaf temperature increased with effective leaf width [fauset et al. 2018]. However, leaf dissection such as lobes, leaf elongation and serration, enhancing a smaller characteristic dimension, thus facilitate efficient heat dissipation, and are evidently characteristics of sun exposed leaves than shaded leaves along a vertical gradient [@vogelConvectiveCoolingLow1970; @schueppTansleyReviewNo1993; @sackHowStrongIntracanopy2006]. Other morphological characteristics such as increased petiole length, characteristic of leaves growing in high-light conditions and in exposed canopy positions, also contribute to heat dissapation [Niinemetes]

@@

##### *Leaf lifespan* [Elsa: phenology more broadly? - leaf lifespan, fruiting, flowering links to temperature and thermal gradients?]
*Leaf lifespan generally increases with LMA. LMA and leaflifespan is greater with available light, and largely depends on cost of carbon investment in a leaf, which varies across species [**double check**] Leaf turnover tends to be greater in the sun (in the tropics), so of your standing stock of leaves (at least in evergreen species) sun leaves should be younger than shade leaves. This is bound to affect metabolism, and may affect other aspects of their performance.*

@@

Across ontogeny, canopy and understory trees of same species in tropical, temperate, and conifer forests show significant differences in leaf traits, however, the difference has been largely attributed to changes in light conditions rather than ontogeny [Rosendaal et al. 2006], therefore maintaing similar responses to changes in light with increasing height such as in within-species. Under constant light conditions of open-grown canopy and understory, for example in large gaps, evidence shows that canopy trees showed significantly greater LMA, photosynthetic capcity per area, and stomatal conductance compared to saplings, attributing the differences in traits to ontogeny [Thomas and Winner, 2002; Houter and Pons 2012]. On the basis of phenology, temperate understory trees have been observed to leaf out earlier than canopy due to ontogenetic variation in tree size rather than biophysical conditions [Vitasse et al. 2013], however, increase in spring temperatures has shown to have implications on understory phenology, relative to canopy trees [Zohnner and Renne, 2019]

>>>>

**(paragraph on radiation as a driver)**
Incoming solar radiation is a strong driver of leaf temperature, relative to $T_{air}$, such that even at low maximum $T_{air}$, leaves exposed to high radiation loads can experience elevated $T_{leaf}$, greater than $T_{air}$. On the other hand, under conditions of constant radiation and water availability, increase in $T_{air}$ can induce evaporative cooling [Drake et al. 2018] or under high $T_{air}$, can cause leaf heat stress [Fauset et al. 2018]. Dense canopies that buffer understory tend to absorb and reflect incoming radiation, therefore maintain greater $T_{leaf}$ than $T_{air}$, whereas open-forests, with greater light interception to forest floors, in come cases, exhibit a less pronounced gradient of leaf temperatures, and in other cases maintain greater $T_{leaf}$-$T_{air}$ in the lower canopies [**REF**]

@@

Particularly in dense forests, the upper, outer canopy leaves of taller trees are exposed to dramatically different biophysical conditions than leaves in the canopy interior or understory (Figs. `r fig_NEON_vertical`, `r SIfigs_NEON_all`). Exposed, typical sun leaves are often smaller in size and width, thicker, and hairier, with thicker cuticle and greater lobation than shade leaves. These traits also facilitates heat dissapation [Vogel] in sun leaves compared to shade. Whereas shade leaves increase light capture efficiency by increasing their specific leaf area [SLA] and by investing in chlorophyll on a mass basis. "This enables shade leaves to realize similar light capture, under low light, to sun leaves at a lower biomass investment" [Chazdon et al. 1996, Poorter et al. 2000; @rozendaalPlasticityLeafTraits2006; Mathur et al. 2018], however might not be most suitable in dissapating heat [Vogel]. Leaf traits are critical to maintaining desirable $T_{leaf}$ and, in turn, shaping leaf metabolism (following section) and carbon balance. It is unsuprising, then, that leaf traits vary dramatically across vertical gradients.

@@

Within-individuals in dense forests, canopy positions are exposed to different light and biophysical conditions in the upper, middle and lower canopy. Thus leaf traits vary accordingly. Whereas in open-forests or under consistent canopy light conditions, leaf traits vary on the basis of light exposure. In the former condition, several studies point towards common variations in leaf anatomical, structural and biochemical trait plasticity at a vertical intra-canopy gradient across individual trees [@sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020]. The differences among these traits contribute to leaf heat dissipation strategies and leaf temperature distribution across the gradient [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016] Table `r table_leaf_traits` summarizes relevant traits for thermal sensitivity.

@@

Leaf temperature is generally assumed to increase with leaf area ($LA$) [ Wright et al. 2019; Tserej & Feeley 2021], however effective leaf width is a better predictor of leaf temperature changes and leaf thermal time constant than leaf area [@leighInfluenceLeafSize2017]. With increases in effective leaf width, thermal time constant also increases due to increase in boundary layer thickness and resistance, along with stomatal resistance [@leighInfluenceLeafSize2017; Wright et al. 2019; @fauset_differences_2018]. Therefore, the rate of heat convection per unit area between leaf and air is lower for large leaves and greater for small leaves that are able to dissipate heat more efficiently than large leaves [@leighInfluenceLeafSize2017]. Evidence shows that leaf temperature increased with effective leaf width [fauset et al. 2018]. However, leaf dissection such as lobes, leaf elongation and serration, enhancing a smaller characteristic dimension, thus facilitate efficient heat dissipation, and are evidently characteristics of sun exposed leaves than shaded leaves along a vertical gradient [@vogelConvectiveCoolingLow1970; @schueppTansleyReviewNo1993; @sackHowStrongIntracanopy2006]. Other morphological characteristics in exposed canopy positions such as increased petiole length, cuticle thickness, trichomes, stomatal density also enhance heat dissapation compared to shade leaves [Niinemetes]

@@

##### *Leaf lifespan* [Elsa: phenology more broadly? - leaf lifespan, fruiting, flowering links to temperature and thermal gradients?]
*Leaf lifespan generally increases with LMA. LMA and leaflifespan is greater with available light, and largely depends on the cost of carbon investment in a leaf, which varies across species sun vs. shade leaves [**double check**] Leaf turnover tends to be greater in the sun (in the tropics), so of your standing stock of leaves (at least in evergreen species) sun leaves should be younger than shade leaves. This is bound to affect metabolism, and may affect other aspects of their performance.*

@@

Across ontogeny, canopy and understory trees of same species in tropical, temperate, and conifer forests show significant differences in leaf traits in their biophysical environment, however, the difference has been largely attributed to changes in light conditions rather than ontogeny itself [Rosendaal et al. 2006], therefore maintaing similar responses to changes in light with increasing height, as observed in within-species gradient. Under constant light conditions of open-grown canopy and understory, for example in large gaps, evidence shows that canopy trees showed significantly greater LMA, photosynthetic capcity per area, and stomatal conductance compared to saplings, attributing the differences in traits to ontogeny [Thomas and Winner, 2002; Houter and Pons 2012]. On the basis of phenology, temperate understory trees tended to leaf out earlier than canopy due to ontogenetic variation in tree size rather than biophysical conditions [Vitasse et al. 2013], and, increase in spring temperatures has shown to have implications on understory phenology, relative to canopy trees [Zohnner and Renne, 2019]
